The five pillars of computational reproducibility: Bioinformatics and beyond

### 2023-06-01 manuscript/5pillars.Rmd
more text editing aroung guix

Another potential solution to this is to use a package/environment management system
(e.g.: [Conda](https://conda.org/) or [Guix](https://guix.gnu.org/)).
These allow users to create, manage, and deploy software packages,
dependencies, and environments across different computing platforms @Perkel2023-cv.
Conda was initially developed to solve the problem of package dependency management in Python,
but it now supports many other programming languages, including R, C/C++, Java, and others.
Conda allows researchers to create isolated environments with specific versions of
packages, so users can have different versions of R or Python in different environments
on the same host system.
Conda environments have limitations around portability and consistency, as such environments
are tied to specific operating systems and hardware architecture.
Guix on the other hand, gives system-level reproducibility, but has a smaller community and ecosystem
of supported packages.

Containers are an attempt to resolve some of the downsides of VMs and package managers.
Container images can be thought of as similar to VMs, but are more lightweight as they
share parts of the host operating system.
In the example shown in **Figure 4**, the six containerised applications share the same operating
system, while the three VM applications each involve their own operating system which incurs a
significant performance overhead @noauthor_2021-lt.
Therefore, running workflows in containers incurs only a small reduction in performance as compared
to running directly on the host system @Di_Tommaso2015-pw.
Container images are highly portable because containers include everything needed to run the
application, including the system tools and libraries, ensuring that the environment is consistent
across different systems.
For example, the most popular containerisation system, Docker [@Merkel2014-ek], makes it
possible to run Windows and Linux/Unix containers on any computer with Docker installed, with the
promise of reproducibility.
While there are several alternatives such as Podman and Rkt, Docker remains the most widely used
containerisation system for web development and research.
Due to its early adoption, Docker now has a large community of users, extensive documentation and
a vast collection of pre-built container images called DockerHub.
Docker can fetch images from DockerHub and run them on the host system with just a couple of commands,
and typically within a few minutes.
This accelerates the installation procedure dramatically which is a known bottleneck for bioinformatic
reproducibility [@Mangul2019-ei;@Mangul2019-xp].

@@

@@

It is not sufficient to simply use these tools, they need to be shared as part of the publication
process.
Conda environments are described by a file called `environment.yml` which acts like a recipe for how to
make the environment.
As they are a recipe, these are small files which are easily shared in the project code repository.
Docker images are larger and cannot be shared in the code repository.
They are commonly deposited to the DockerHub repository, but availability of images is dependent on the
commercial viability of Docker Inc, so it isn't guaranteed to survive the 10-year challenge.
A better option for bioinformatics workflows would be to use the BioContainers registry
@Da_Veiga_Leprevost2017-yw, as it is community driven and supported by consortia such as *Elixir* and
*Global Alliance for Genomics and Health* that will underpin its longevity.
Still, these are not considered permanent archival services, so it is advised to deposit the image
used in an article to a suitable persistent long-term archive (discussed below).
The recipe used to build the Docker image, called the `Dockerfile`, is sufficiently small and can be
added to the project source code repository.

@@

@@

We are grateful to the individuals who provided feedback on an earlier draft of the concept including
Dr Aziz Khan (Stanford University),
Dr Sriganesh Srihari (QIMR Berghofer Medical Research Institute),
Dr Barbara Zdrazil (EMBL-EBI) and
Dr Irene López Santiago (EMBL-EBI).

>>>>

Containers are an attempt to resolve some of the downsides of VMs and package managers.
Container images can be thought of as similar to VMs, but are more lightweight as they
share parts of the host operating system [@Nust2020-cr].
In the example shown in **Figure 4**, the six containerised applications share the same operating
system, while the three VM applications each involve their own operating system which incurs a
significant performance overhead @noauthor_2021-lt.
Therefore, running workflows in containers incurs only a small reduction in performance as compared
to running directly on the host system @Di_Tommaso2015-pw.
Container images are highly portable because containers include everything needed to run the
application, including the system tools and libraries, ensuring that the environment is consistent
across different systems.
For example, the most popular containerisation system, Docker [@Merkel2014-ek], makes it
possible to run Windows and Linux/Unix containers on any computer with Docker installed, with the
promise of reproducibility.
While there are several alternatives such as Podman and Rkt, Docker remains the most widely used
containerisation system for web development and research.
Due to its early adoption, Docker now has a large community of users, extensive documentation and
a vast collection of pre-built container images called DockerHub.
Docker can fetch images from DockerHub and run them on the host system with just a couple of commands,
and typically within a few minutes.
This accelerates the installation procedure dramatically which is a known bottleneck for bioinformatic
reproducibility [@Mangul2019-ei;@Mangul2019-xp].

@@

Another potential solution to this problem is to use a package/environment management system such as
[Conda](https://conda.org/) or [Guix](https://guix.gnu.org/).
These allow users to create, manage, and deploy software packages,
dependencies, and environments across different computing platforms @Perkel2023-cv.
Conda was initially developed to solve the problem of package dependency management in Python,
but it now supports many other programming languages, including R, C/C++, Java, and others.
Conda allows researchers to create isolated environments with specific versions of
packages, so users can have different versions of R or Python in different environments
on the same host system.
Conda environments have limitations around portability and consistency, as such environments
are tied to specific operating systems and hardware architecture.

Guix is described as a "functional package manager" and has the unique benefit of giving
system-level build reproducibility [@Vallet2022-no].
This is a notable advantage, as Docker images are not guaranteed to build reproducibly in the future
due to link decay.
Moreover, Guix can build and output environments as Docker-compatible images which solves the
provenance problem of such environments [@GNU_Project_Volunteers2022-bt].
These features are being used in the bioinformatics sphere to create highly reproducible pipelines
that work equally well on personal as well as shared computer systems @Wurmus2018-lv.
Although Guix has some exciting functionality, there remains a relative lack of step-by-step guides and
tutorials, illustrated by the complete absence of published Guix protocols in protocols.io, as compared
to 13 and 12 related to "Conda" and "Docker" respectively (as at 30th May 2023).

@@

It is not sufficient to simply use these tools, they need to be shared as part of the publication
process.
Conda environments are described by a file called `environment.yml` which acts like a recipe for how to
make the environment.
Similarly, Guix environments rely on two files: `channels.scm` and `manifest.scm`.
Docker images are built using a `Dockerfile` instruction set.
These are small files which are easily shared in the project code repository.
To enable rapid reproduction, making Docker images available is recommended.
As these are often very large they cannot be shared in the code repository; rather
they are commonly deposited to a Docker image registry such as DockerHub.
One must keep in mind that availability of images is dependent on the
commercial viability of Docker Inc, so it isn't guaranteed to survive the 10-year challenge.
To ensure long-term availability, it would be advised to deposit images to the BioContainers registry
@Da_Veiga_Leprevost2017-yw, as it is community driven and supported by consortia such as *Elixir* and
*Global Alliance for Genomics and Health* that will underpin its longevity.
Still, these are not considered permanent archival services, so it is advised to deposit the image
used in an article to a suitable persistent long-term archive (discussed below).
The recipe used to build the Docker image, called the `Dockerfile`, is sufficiently small and can be
added to the project source code repository.

@@

## Conclusion

Reproducibility is an essential feature of reliable computational research yet current practices
are falling short of this goal.
Our recommendations for workflow-based research articles are summarised as:

0. Encapsulate the entire workflow with end-to-end code, executable with a master script.

1. Share the code repository and deposit a snapshot to an archive.

2. Share a controlled environment; options include a Docker or virtual machine image or
configuration for Conda or Guix.

3. Use literate scripting tools like R Markdown or Jupyter notebooks to generate all outputs,
such as charts and tables.

4. Document the workflow sufficiently for reproduction, especially the repository `README`.

5. Share research data and metadata according to FAIR principles.

@@

We are grateful to individuals who provided feedback on an earlier draft of the concept including
Dr Aziz Khan (Stanford University),
Dr Sriganesh Srihari (QIMR Berghofer Medical Research Institute),
Dr Barbara Zdrazil (EMBL-EBI) and
Dr Irene López Santiago (EMBL-EBI).
